A Component-Based Sensor Network for \\ Environmental Monitoring

### 2009-03-13 docs/icccn2009/paper.tex
Edited section on DSP
git-svn-id: svn://localhost/netbeams/trunk@373 559edde2-a4d4-44e0-96fc-4035e40d7396

Sensor networks, especially of environmental sensors, are normally deployed
over distributed locations where communication infrastructure are not always 
reliable or, at least, not countinuously available. Additionally, different
sensors and communication protocols might be available in throughout the network.

As the network grows in size and complexity scalability, reliabilty and maintainability
become critical issues. Additional components 
might be added to the network, the network configuration might be altered or 
new versions of components might be deployed. The sensor network must be able to 
support such adjustments without significant overhead for network administrators and
the components in the network.

From system development and integration point of view, interfaces among components and
between components and the network should be simple. Complex interaction can be 
constructed on top of these simple interfaces ( a good example of this strategy is the 
world-wide-web).

In the following we discuss how our proposed architechure addresses those issues.

@@

Figure \ref{FIG_DSP} depicts the basic architecture of the
DSP. We refer to the plug-and-play modules as DSP Components. DSP
Components are self-contained modules that can be added and removed to
a DSP installation at runtime. The basic communication paradigm within
a DSP is a message. DSP Components can exchange messages of different
types. Actually, all interactions between DSP Components are implemented
through messages yielding simple but powerful communication instructure.
We distinguish between Data Producers (DP) and Data Consumers
(DC) within the scope of a DSP. DP and DC can be seen as roles that a
DSP Component may have. It is possible for a DSP Component to act in
the role of DC and DP at the same time. Messages in DSPs are routed based 
on rules. Rules can be statically or dynamically configured. Rules can use 
different source information when routing a message: producers or consumers of the message, 
the message itself or specific configuration provided by the administrators.

\begin{figure}
\centering
\epsfig{file=dsp-platform, width=7cm}
\caption{\label{FIG_DSP} Data Sensor Platform (DSP).}
\end{figure}

The DSP is responsible for routing messages between DPs and DCs. It is
important to note that this routing only happens withing a
DSP. Following the paradigm of a micro-kernel approach, the DSP has no
notion of remote communication or sensor-specific details. All this
knowledge is embodied in DSP Components.

@@

A DSP Component encapsulates a particular functionality. We
distinguish between those components that produce data (DP) and those
that consume data (DC). An example for a DSP Component acting in the
role of a DP is a sensor module. The sensor module reads data from the
physical sensor and converts the data into a message that is forwarded
to the DSP. Because this module generates data with respect to the
DSP, it is a data producer. An example for a data consumer is a
converter module that converts the internal, DSP-specific data format,
suitable for consumption by external applications. One such common
format used in the domain of environmental monitoring is OpenDAP
\cite{opendap01}. Since the converter module accepts messages from the DSP, it
acts in the role of a data consumer. Finally, a DSP Component
implementing persistency is an example of a module that acts in the
roles of both data producer and consumer. It accepts messages from the
DSP and stores them in a database. At a later time, it may act as a
data producer by resending previously stored messages.

@@

\begin{code}
interface DSPComponent
{
   public String getComponentType();
   public void initComponent(...);
   public void startComponent();
   public void stopComponent();
   public void deliver(DSPMessage msg);
   public Message deliverWithReply(DSPMessage msg);
   // ...
}
\end{code}

By implementing the above Java interface, the resuling code
effectively becomes a DSP Component that can be deployed at runtime. A
component is identified by its type. Only one component of a certain type per DSP 
is permissible. A type identifier much be unique across all DSP Components in DSP. 
Upon deploying a DSP Component, an initialization function is called (line 4). Once a
component is initialized, it may be started (line 5) and stopped (line
6) multiple time. Whenever the DSP wishes to pass a message (that was
generated by some DP), it invokes the \texttt{deliver()} method (ine
7). The message structure will be explained in the following section.
One important feature of the DSP is the ability to redeploy (maybe with a new version) at 
runtime. This allows DSPs to be updated remotely. This process is managed through
messages sent to a special DSP Component which represents the DSP itself.

\subsection{Message}

Message passing is the main paradigm for exchanging information within
a DSP. After producing data, DP component wraps that data in the 
body of a message, so that DC components can receive the message and consume 
the data on the message's body. When it comes to passing messages to a remote
DSP instane, messages are serialized in XML \cite{xml2000} and automatically 
deserialized for DC located in the remote DSP. In order to serialize and
deserialize DSP messages, the DSP infrastructure uses the XML Schema 
standard \cite{xml-schema2004} to generate and validate the DSP message. Figure
\ref{FIG_DSP_MESSAGE} defines the structure of a general DSP Message.

\begin{figure}[!htb]
 \centering
 \epsfig{file=dsp-message, width=9cm, height=4cm}
 \caption{\label{FIG_DSP_MESSAGE} Structure of a DSP Message.}
\end{figure}

The above XML Schema is defined for an ``Abstract Message'' type, whose
structure serves as a base schema for all different types of messages. Any
message will contain two basic attributes: ``messageID'' and ``ContentType''. 
The former uniquely identifies a message and the latter the data in the body of
the message. Complementary identification information such as the message 
producer and consumer are located in the header of the message, which are
defined by the schema type ``ComponenIdentifier''. A ``ComponentIdentifier''
carries out information about the component name and its physical host address.
Finally, the body of the message may contain any data type defined by the component 
developer, and it can be interpreted as the DSP payload.

Although XML schema is the de-facto standard to bind and validate xml data to
its type, the architectural infrastructure of DSP uses an XML Data Bindining
Technology \cite{xml-dbind} called JAXB \cite{xml-jaxb}. It is used to generate
a type safe API for handling in-memory object graphs from the XML instances of 
DSP Messages and the content of its data payload. Moreover, taking advantage 
of the data type polimorphirsm of XML schema data types - concept borrowed from 
Object-Oriented Design \cite{Rumbaugh1991} - we defined different types of 
messages in order to allow specific DP and DC interpret the payload of any 
given message solely based on their runtime type, as follows:

\begin{itemize}
  \item Action Message;
  \item Event Message;
  \item Create Message;
  \item Insert Message;
  \item Update Message;
  \item Delete Message;
  \item Query Message;
  \item Measure Message.
\end{itemize}

Not only the message structure is defined by XML schemas, but also the body
content is and may be of any user-defined type. Therefore, the data
defined to be sent from any given DP to any DC can be also in XML format or
in-memory. An example illustrating a DSP message is given in the YSI section.

\subsection{Broker}

As soon as a DP produces data and wraps it up in a DSP Message, the DP only
interact with the DSP platform via the DSP Broker by making the call
``Broker.deliver(DSP Message)''. In this way, DSP broker is responsible for
delivering the produced message to DSP consumers DCs, as well as verifying
whether the delivery of the message is going to be done locally or remotely.

\begin{itemize}
 \item Local delivery of messages: it uses local object reference to deliver the
DSP message;
 \item Remote delivery of messages: it uses any DSP component that is
specialized in remote delivery defined as ``Gateway DSP Component''.
\end{itemize}

The selection of the DCs for a given message to where the Broker will deliver
messages or use a gateway component is determined by an object called Matcher.

\subsection{Matcher}

The Matcher is an object responsible to find the possible set of DCs for a given
DSP message, and its ``deterministic intelligence'' relies on matching rules
defined by the DSP administrator. In this way, a set of matching rules is define
in XML and stored at the DSP run-time directory, where each matching rule is
uniquely identified by a rule ID and determines the following (see
figure \ref{FIG_MATCHER_RULE}):

@@

\begin{itemize}
 \item Match Criteria: given the DSP message delivered to the Broker, which
component(s) produced that message and in which node? It defines the name of
the component type of the producer and its physical location (IP address);
 \item Match Target: given the DSP message, which component(s) are going
to consume the message, and in which node? It defines the name of the component
type of the consumer, and optionally defines a gateway component.
\end{itemize}

\subsection{Matcher replies to Broker}

When analyzing the matching rules, the Matcher only selects the rules that
satisfy the Matching Criteria and forwards them back to the Broker. Once the
broker has the matching rules, it determines whether the delivery is done
directly to a component defined on the Match Target, or if the message is
delivered to the optional gateway. We have defined special gateway DSP
components that is used to transport the DSP Messages, serialized in
XML, to any remote DSP instance.

\subsection{OSGi}

OSGi lends itself naturally to implement the DSP.

DSP can be implemented in different ways, but OSGi is a natural match.
DSP Components are mapped to bundles. DSP itself is a bundle.

>>>>

Sensor networks, especially of environmental sensors, are normally
deployed over a geographically wide area where the communication
infrastructure is not always reliable or, at least, not countinuously
available. As the sensor network grows in size and complexity
scalability, reliabilty and maintainability become critical issues.
Additional components might be added to the network, sensor
configuration might be altered, or new versions of components might be
deployed. The sensor network must be able to support such adjustments
without significant overhead for maintainers.  From system
development and integration point of view, interfaces among components
and between components and the network should be simple. Complex
interaction can be constructed on top of these simple interfaces.  In
the following we discuss how our proposed architechure addresses those
issues.

@@

Figure \ref{FIG_DSP} depicts the basic architecture of the DSP. We
refer to the plug-and-play modules as DSP Components. DSP Components
are self-contained modules that can be added and removed to a DSP
installation at runtime. The basic communication paradigm within a DSP
is a message. DSP Components can exchange messages of different types.
Actually, all interactions between DSP Components are implemented
through messages yielding simple but powerful communication
instructure.  We distinguish between \emph{Data Producers} (DP) and
\emph{Data Consumers} (DC) within the scope of a DSP. DP and DC can be
seen as roles that a DSP Component may have. It is possible for a DSP
Component to act in the role of DC and DP at the same time. Messages
in DSPs are routed based on rules. Rules can be configured statically
or dynamically. Rules for routing messages can be based on different
information: producers or consumers of the message, the message
itself, or specific configuration parameters provided externally.

\begin{figure}
\centering
\epsfig{file=dsp, width=7cm}
\caption{\label{FIG_DSP} Data Sensor Platform (DSP).}
\end{figure}

The DSP is responsible for routing messages between DPs and DCs. It is
important to note that this routing only happens within a
DSP. Following the paradigm of a micro-kernel approach, the DSP has no
notion of remote communication or sensor-specific details. All this
knowledge is embodied in special purpose DSP Components.

@@

A DSP Component encapsulates a particular functionality. We
distinguish between those components that produce data (DP) and those
that consume data (DC). An example for a DSP Component acting in the
role of a DP is a module that interfaces with a physical sensor. Such
a module reads data from the sensor and converts the data into a
message that is forwarded to the DSP. Because this module generates
data with respect to the DSP, it is a data producer. An example for a
data consumer is a converter module that converts the internal,
DSP-specific data format to a format suitable for consumption by
external applications. One such common format used in the domain of
environmental monitoring is OpenDAP \cite{opendap01}.  Since the
converter module accepts messages from the DSP, it acts in the role of
a data consumer. Finally, a DSP Component implementing persistency is
an example of a module that acts in the roles of both data producer
and consumer. It accepts messages from the DSP and stores them in a
database. At a later time, it may act as a data producer by resending
previously stored messages.

@@

\begin{code}
interface DSPComponent
{
   public String getComponentType();
   public void initComponent(...);
   public void startComponent();
   public void stopComponent();
   public void deliver(DSPMessage msg);
   // ...
}
\end{code}

By implementing the above Java interface, the resuling code
effectively becomes a DSP Component that can be deployed on a DSP at
runtime. A component is identified by its type (line 3). Only one
component of a certain type per DSP is permissible. A type identifier
must be unique across all DSP Components in a DSP.  Upon deploying a
component, an initialization function is called (line 4). Once a
component is initialized, it may be started (line 5) and stopped (line
6) multiple time. Whenever the DSP wishes to pass a message (that was
generated by some DP), it invokes the \texttt{deliver()} method (line
7). The message structure will be explained in the following section.
One important feature of the DSP is the ability to redeploy (maybe
with a new version) at runtime. This allows DSPs to be updated
remotely. This process is managed through messages sent to a special
DSP Component that is able to manage its local DSP.

\subsection{Message Structure}

Message passing is the main paradigm for exchanging information within
a DSP. After producing data, a DP wraps that data in the body
of a message, so that DC components can receive the message and
consume the data contained in the message's body. When it comes to passing
messages to a remote DSP instance, messages are serialized in XML
\cite{xml2000} and automatically deserialized for DC located in the
remote DSP. In order to serialize and deserialize DSP messages, the
DSP infrastructure uses the XML Schema standard \cite{xml-schema2004}
to generate and validate the DSP message. Figure \ref{FIG_DSP_MESSAGE}
defines the structure of a general DSP message.

\begin{figure}[!htb]
 \centering
 \epsfig{file=dsp-message, width=9cm, height=4cm}
 \caption{\label{FIG_DSP_MESSAGE} Structure of a DSP message.}
\end{figure}

The basic message structure is defined by an \texttt{AbstractMessage}
type, whose structure serves as a base schema for all different types
of messages. Any message will contain two basic attributes:
\texttt{messageID} and \texttt{ContentType}.  The former uniquely
identifies a message and the latter the data in the body of the
message. Complementary identification information such as the message
producer and consumer are located in the header of the message, which
are defined by the schema type \texttt{ComponenIdentifier}. A
\texttt{ComponentIdentifier} carries out information about the
component name and its physical host address.  Finally, the body of
the message represents the payload and may contain any data type
defined by the component developer.

Although XML schema is the de-facto standard to bind and validate XML
data to its type, the architectural infrastructure of DSP uses an XML
data binding technology \cite{xml-dbind} called JAXB \cite{xml-jaxb}.
It is used to generate a type safe API for handling in-memory object
graphs from the XML instances of DSP Messages and the content of its
data payload. Moreover, taking advantage of the data type polymorphism
of XML schema data types -- a concept borrowed from object oriented
principles -- we defined different categories of messages in order to
reflect the nature of the message content. Among those categories are
query and update messages for configuring various properties supported
by DSP Components and measurement messages that actually contain
sensor readings.

The payload of a message is itself described via an XML schema in
order to integrate it seemingly with the general message structure. A
developer of a DSP Component can define a schema that represents the
type of payload processed by that component. An example illustrating a
DSP message is given when we discuss the YSI Sonde Component.

\subsection{DSP Broker}

In order to keep the API symmetrical, the DSP is itself implemented as
a DSP Component, the so-called DSP Broker. While technically the DSP
Broker has the same API as any component, it serves a special role and
cannot be removed. When a DP sends a message, it actually
invokes the \texttt{deliver()} method of the DSP Broker. The broker is
responsible for routing the message to one or several DCs. It
distinguishes between local and remote destinations:

\begin{itemize}
\item Local delivery of messages: it uses local object reference to
  deliver the DSP message.
\item Remote delivery of messages: it uses a DSP Component marked as a
  default gateway to handle remote message delivery.
\end{itemize}

The selection of the DCs for a given message to where the broker will
deliver messages to is determined by a matcher.  The matcher is
responsible to find the possible set of DCs for a given DSP message,
and it relies on matching rules defined statically or at runtim. In
this way, a set of matching rules is defined in XML and stored at the
DSP run-time directory, where each matching rule is uniquely
identified by a rule ID and determines the following (see Figure
\ref{FIG_MATCHER_RULE}):

@@

\begin{itemize}
\item The \emph{match criteria} defines the name of the component type
  of the producer and its physical location (IP address);
\item The \emph{match target} defines the name of the component type
  of the consumer, and optionally defines a gateway component.
\end{itemize}

When analyzing the matching rules, the matcher only selects the rules
that satisfy the matching criteria and forwards them back to the
broker. Once the broker has the matching rules, it determines whether
the delivery is done directly to a component defined on the match
target, or if the message is delivered to the optional gateway. We
have defined special gateway DSP components that is used to transport
the DSP Messages, serialized in XML, to any remote DSP instance.
